---
title: "COVIDEDA (some W203 lab 2 data explorations)"
author: "Tilman Bayer, Kieran Berton, Ruilan Liu, Toby Petty"
output:
  html_document:
    df_print: paged
---
Reusing parts of [Kevin Martin's "COVID Data Chomper (CDC)"](https://ucbischool.slack.com/archives/C01HVT8V2BV/p1616362670110600) ("This tool will grab data from the New York Times Covid case tracker dataset on github and help you chomp it up into a form that is more useful to you. Feel free to modify and supplement this file as desired.")
1
## Import libraries

```{r setup, include=FALSE}
## Import libraries
library("stargazer")
library("data.table")
library("ggplot2")
library("dplyr")
library("lmtest")
library("psych")
library("patchwork")
library("leaps")
library("ggpubr") # run 'install.packages("ggpubr")' once manually if necessary
```

## Ingest data from NYT COVID Database 

(Downloaded on March 27, 2021 via `$ curl -O https://raw.githubusercontent.com/nytimes/covid-19-data/eab41bf87e3fed34ff31f30b2a1cc1ec73377eff/us-states.csv`. Dataset documentation: https://github.com/nytimes/covid-19-data )
```{r get_nyt_data}
NYT_Data <- fread("../data/raw/nytimes/us-states.csv")

#convert dates to date
NYT_Data[,date:=as.Date(date)]

NYT_Data_increments <- NYT_Data %>% 
  group_by(state) %>% 
    mutate(daily_cases = cases-lag(cases, default = cases[1]))

NYT_Data_increments <- NYT_Data_increments %>% 
  group_by(state) %>% 
    mutate(daily_deaths = deaths-lag(deaths, default = deaths[1]))

#take total deaths from final date in dataset
static_date <- max(NYT_Data_increments$date)
cat(as.character(static_date), "is the last date in the dataset")

#exclude non-state territories and non-final dates
NYT_covid <- NYT_Data_increments %>%
  filter(date == static_date,
    state != "Virgin Islands",
    state != "Guam",
    state != "Northern Mariana Islands",
    state != "District of Columbia",
    state != "Puerto Rico"
  ) %>% 
  subset(select = -c(daily_cases,daily_deaths))
```


## Ingest 2019 American Community Survey data 

(Downloaded on March 27, 2021 from https://data.census.gov/cedsci/table?q=ACS&g=0100000US.04000.001&tid=ACSDP1Y2019.DP05&moe=false&hidePreview=true )
```{r get_acs_data}

# ACS population data
ACS_pop <- fread("../data/raw/ACS/ACSDP1Y2019.DP05_data_with_overlays_2021-03-25T141408.csv",skip=1)

demographic <- select(ACS_pop, 
                      "Geographic Area Name", 
                      "Estimate!!SEX AND AGE!!Total population", 
                      "Estimate!!SEX AND AGE!!Total population!!65 to 74 years",
                      "Estimate!!SEX AND AGE!!Total population!!75 to 84 years",
                      "Estimate!!SEX AND AGE!!Total population!!85 years and over") %>%
               rename(
                 state = "Geographic Area Name",
                 total_population = "Estimate!!SEX AND AGE!!Total population",
                 pop_age_65 = "Estimate!!SEX AND AGE!!Total population!!65 to 74 years",
                 pop_age_75 = "Estimate!!SEX AND AGE!!Total population!!75 to 84 years",
                 pop_age_85 = "Estimate!!SEX AND AGE!!Total population!!85 years and over") %>%
               mutate(
                 pop_age_75_plus = pop_age_75 + pop_age_85,
                 pop_age_65_plus = pop_age_65 + pop_age_75 + pop_age_85)



# ACS poverty data
ACS_poverty <- fread("../data/raw/ACS/ACSST1Y2019.S1701_2021-03-29T143251/ACSST1Y2019.S1701_data_with_overlays_2021-03-25T140047.csv",skip=1)

poverty <- select(ACS_poverty, 
                  "Geographic Area Name",
                  "Estimate!!Below poverty level!!Population for whom poverty status is determined!!AGE!!65 years and over",
                  "Estimate!!Below poverty level!!Population for whom poverty status is determined!!EDUCATIONAL ATTAINMENT!!Population 25 years and over!!Some college, associate's degree",
                  "Estimate!!Below poverty level!!Population for whom poverty status is determined!!EDUCATIONAL ATTAINMENT!!Population 25 years and over!!Bachelor's degree or higher",
                  "Estimate!!Percent below poverty level!!Population for whom poverty status is determined",
                  "Estimate!!Total!!Population for whom poverty status is determined") %>%
  rename(
    state = "Geographic Area Name",
    poverty_65_plus = "Estimate!!Below poverty level!!Population for whom poverty status is determined!!AGE!!65 years and over",
    poverty_some_college = "Estimate!!Below poverty level!!Population for whom poverty status is determined!!EDUCATIONAL ATTAINMENT!!Population 25 years and over!!Some college, associate's degree",
    poverty_bachelor = "Estimate!!Below poverty level!!Population for whom poverty status is determined!!EDUCATIONAL ATTAINMENT!!Population 25 years and over!!Bachelor's degree or higher",
    population_in_poverty = "Estimate!!Total!!Population for whom poverty status is determined",
    percent_in_poverty = "Estimate!!Percent below poverty level!!Population for whom poverty status is determined") %>%
  mutate(
    poverty_higher_edu = poverty_some_college + poverty_bachelor)



# ACS education data
ACS_edu <- fread("../data/raw/ACS/ACSST1Y2019.S1501_2021-03-30T021803/ACSST1Y2019.S1501_data_with_overlays_2021-03-28T203228.csv",skip=1)

education <- select(ACS_edu,
                    "Geographic Area Name",
                    "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 18 to 24 years!!Some college or associate's degree",
                    "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 18 to 24 years!!Bachelor's degree or higher",
                    "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 25 years and over!!Some college, no degree",
                    "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 25 years and over!!Associate's degree",
                    "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 25 years and over!!Bachelor's degree or higher",
                    "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 65 years and over!!Bachelor's degree or higher",
                    "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 18 to 24 years!!High school graduate (includes equivalency)",
                    "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 25 years and over!!High school graduate (includes equivalency)") %>%
  rename(
    state = 'Geographic Area Name',
    edu_assoc_under_24 = "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 18 to 24 years!!Some college or associate's degree",
    edu_bachelor_under_24 = "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 18 to 24 years!!Bachelor's degree or higher",
    edu_nodegree_over_25 = "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 25 years and over!!Some college, no degree",
    edu_assoc_over_25 = "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 25 years and over!!Associate's degree",
    edu_bachelor_over_25 = "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 25 years and over!!Bachelor's degree or higher",
    edu_bachelor_over_65 = "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 65 years and over!!Bachelor's degree or higher",
    edu_hs_under_24 = "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 18 to 24 years!!High school graduate (includes equivalency)",
    edu_hs_over_25 = "Estimate!!Total!!AGE BY EDUCATIONAL ATTAINMENT!!Population 25 years and over!!High school graduate (includes equivalency)") %>%
  mutate(
    edu_higher_edu_18_plus = edu_assoc_under_24 + edu_bachelor_under_24 + edu_nodegree_over_25 + edu_assoc_over_25 + edu_bachelor_over_25,
    edu_higher_edu_25_plus = edu_nodegree_over_25 + edu_assoc_over_25 + edu_bachelor_over_25,
    edu_hs_18_plus = edu_hs_under_24 + edu_hs_over_25,
    edu_hs_or_higher_18_plus = edu_hs_under_24 + edu_hs_over_25 + edu_higher_edu_18_plus,
    edu_hs_or_higher_25_plus = edu_hs_over_25 + edu_higher_edu_25_plus)



# ACS housing data
ACS_housing <- fread("../data/raw/ACS/ACSDP1Y2019.DP04_2021-03-30T022519/ACSDP1Y2019.DP04_data_with_overlays_2021-03-29T141407.csv",skip=1)

housing <- select(ACS_housing,
                  "Geographic Area Name",
                  "Estimate!!HOUSING OCCUPANCY!!Total housing units",
                  "Estimate!!OCCUPANTS PER ROOM!!Occupied housing units!!1.00 or less",
                  "Estimate!!OCCUPANTS PER ROOM!!Occupied housing units!!1.01 to 1.50",
                  "Estimate!!OCCUPANTS PER ROOM!!Occupied housing units!!1.51 or more") %>%
            rename(
              state = "Geographic Area Name",
              total_housing_units = "Estimate!!HOUSING OCCUPANCY!!Total housing units",
              less_than_1person_room = "Estimate!!OCCUPANTS PER ROOM!!Occupied housing units!!1.00 or less",
              more_than_1person_room = "Estimate!!OCCUPANTS PER ROOM!!Occupied housing units!!1.01 to 1.50",
              more_than_1.5person_room = "Estimate!!OCCUPANTS PER ROOM!!Occupied housing units!!1.51 or more") %>%
            mutate(
              one_plus_person_room = more_than_1person_room + more_than_1.5person_room )



# ACS internet data
ACS_internet <- fread("../data/raw/ACS/ACSST1Y2019.S2801_2021-04-08T171225/ACSST1Y2019.S2801_data_with_overlays_2021-04-08T171101.csv",skip=1)

internet <- select(ACS_internet,
                   "Geographic Area Name",
                   "Estimate!!Total!!Total households",
                   "Estimate!!Total!!Total households!!TYPES OF COMPUTER!!Has one or more types of computing devices:",
                   "Estimate!!Total!!Total households!!TYPES OF COMPUTER!!No computer",
                   "Estimate!!Total!!Total households!!TYPE OF INTERNET SUBSCRIPTIONS!!With an Internet subscription:",
                   "Estimate!!Total!!Total households!!TYPE OF INTERNET SUBSCRIPTIONS!!Without an Internet subscription") %>%
            rename(
              state = "Geographic Area Name",
              total_households = "Estimate!!Total!!Total households",
              with_computer = "Estimate!!Total!!Total households!!TYPES OF COMPUTER!!Has one or more types of computing devices:",
              wo_computer = "Estimate!!Total!!Total households!!TYPES OF COMPUTER!!No computer",
              with_internet = "Estimate!!Total!!Total households!!TYPE OF INTERNET SUBSCRIPTIONS!!With an Internet subscription:",
              wo_internet = "Estimate!!Total!!Total households!!TYPE OF INTERNET SUBSCRIPTIONS!!Without an Internet subscription") %>%
            mutate(
              wo_computer_pct = wo_computer/total_households,
              wo_internet_pct = wo_internet/total_households)
```

## Land Area Data
```{r import land area data}
# US Area data
CB_area <- fread("../data/raw/LandArea/State Area Measurements and Internal Point Coordinates.csv")

landarea <- select(CB_area,State,LandArea_Sq.Mi.) %>%
  rename(state = State,
         landarea = LandArea_Sq.Mi.)
```

```{r}
pop_and_poverty = merge(poverty,demographic)
write.csv(pop_and_poverty,"../data/processed/pop_and_poverty.csv", row.names = FALSE)

```

## Augment NYT data with ACS statistics and Land Area Data
```{r augment_with_acs}
#combine data on covid cases/deaths each day with data on state populations
data_cleaned <- merge(NYT_covid, demographic)
data_cleaned <- merge(data_cleaned,poverty)
data_cleaned <- merge(data_cleaned,education)
data_cleaned <- merge(data_cleaned,housing)
data_cleaned <- merge(data_cleaned,internet)
data_cleaned <- merge(data_cleaned,landarea)
data_cleaned <- data_cleaned %>%
    mutate(
      casesperlakh = 100000 * cases / total_population,
      deathsperlakh = 100000 * deaths / total_population,
      pop_den = total_population / landarea,
      log_pop_den = log10(pop_den),
      house_per_capita = total_housing_units / total_population,
      hs_edu_18_plus_percent = edu_hs_18_plus / total_population,
      hs_edu_25_plus_percent = edu_hs_over_25 / total_population,
      hs_or_higher_18_plus_percent = edu_hs_or_higher_18_plus / total_population,
      hs_or_higher_25_plus_percent = edu_hs_or_higher_25_plus / total_population,
      higher_edu_18_plus_per_capita = edu_higher_edu_18_plus/total_population,
      higher_edu_25_plus_per_capita = edu_higher_edu_25_plus/total_population,
      pct_age_65_plus = pop_age_65_plus/total_population,
      pct_age_75_plus = pop_age_75_plus/total_population,
      pct_age_85_plus = pop_age_85/total_population,
      )

write.csv(data_cleaned,"../data/processed/data_cleaned.csv", row.names = FALSE)
```


# Histograms of variables in the data_cleaned table
```{r message = FALSE, error=FALSE,echo=FALSE}
sapply(data_cleaned, is.numeric)

# plot series of histograms, excluding variables not normalized by state size
for_hist <- subset(data_cleaned,select = 
                     -c(state,date,fips,cases,deaths,
                        pop_age_65, pop_age_75, pop_age_85,
                        pop_age_65_plus, pop_age_75_plus, 
                        poverty_65_plus, poverty_some_college, poverty_bachelor,
                        poverty_higher_edu, edu_assoc_under_24,
                        edu_bachelor_under_24, edu_nodegree_over_25,
                        edu_assoc_over_25, edu_bachelor_over_25,
                        edu_bachelor_over_65, edu_higher_edu_18_plus,
                        edu_higher_edu_25_plus,
                        edu_hs_under_24,
                        edu_hs_over_25,
                        edu_hs_18_plus,
                        edu_hs_or_higher_18_plus,
                        edu_hs_or_higher_25_plus,
                        total_housing_units,
                        less_than_1person_room, more_than_1person_room,
                        more_than_1.5person_room, one_plus_person_room,
                        total_households, 
                        with_computer, wo_computer, with_internet, wo_internet,
                        landarea
                        ))

par(mfrow = c(2, 2))
lapply(seq(for_hist), function(x)
  hist(x=for_hist[[x]],xlab=names(for_hist)[x], main=paste("hist", names(for_hist)[x])))

```


# scatterplots to decide transformation
```{r}

par(mfrow = c(2, 2))

loop.vector <- 1:ncol(for_hist)

for (i in loop.vector) {
  
  x <- for_hist[,i]
  y <- for_hist$deathsperlakh
  plot(x,y,main = paste("deathsperlakh vs.",names(for_hist)[i]),xlab=names(for_hist)[i],ylab="deathsperlakh")

}

```



# Model evaluation 

## Define a model 

### Try subset selection with both pop_den and log_pop_den included
```{r subset selection including both pop_den and log_pop_den}
data_cleaned_model_selection_with_lpd <- data_cleaned %>% select(
  percent_in_poverty,
  wo_internet_pct,
  deathsperlakh,
  pop_den,
  log_pop_den,
  house_per_capita,
  hs_or_higher_25_plus_percent,
  higher_edu_18_plus_per_capita,
  pct_age_85_plus
)

# Adapted from http://www.sthda.com/english/articles/37-model-selection-essentials-in-r/155-best-subsets-regression-essentials-in-r/#computing-best-subsets-regression 
# (but without their use of BIC and CP):
models <- regsubsets(deathsperlakh~., data = data_cleaned_model_selection_with_lpd,force.in = "percent_in_poverty")
summary(models)

res.sum <- summary(models)
Adj.R2 <- which.max(res.sum$adjr2)
print(res.sum$adjr2)
cat('Optimal number of predictors to minimize adjusted R^2:', Adj.R2)
max_num_predictors <- length(res.sum$adjr2)

get_names <- coef(models, 1:sum(lengths(res.sum$adjr2)))

```


### Try subset selection with only log_pop_den included

```{r subset selection including only log_pop_den}
data_cleaned_model_selection_with_only_lpd <- data_cleaned %>% select(
  percent_in_poverty,
  wo_internet_pct,
  deathsperlakh,
  # pop_den,
  log_pop_den,
  house_per_capita,
  hs_or_higher_25_plus_percent,
  higher_edu_18_plus_per_capita,
  pct_age_85_plus
)

# Adapted from http://www.sthda.com/english/articles/37-model-selection-essentials-in-r/155-best-subsets-regression-essentials-in-r/#computing-best-subsets-regression 
# (but without their use of BIC and CP):
models <- regsubsets(deathsperlakh~., data = data_cleaned_model_selection_with_only_lpd,force.in = "percent_in_poverty")
summary(models)

res.sum <- summary(models)
Adj.R2 <- which.max(res.sum$adjr2)
print(res.sum$adjr2)
cat('Optimal number of predictors to minimize adjusted R^2:', Adj.R2)
max_num_predictors <- length(res.sum$adjr2)

get_names <- coef(models, 1:sum(lengths(res.sum$adjr2)))

```


### (Final choice:) Subset selection with only pop_den included

```{r subset selection without log_pop_den (final version)}
data_cleaned_model_selection <- data_cleaned %>% select(
  percent_in_poverty,
  wo_internet_pct,
  deathsperlakh,
  pop_den,
  house_per_capita,
  hs_or_higher_25_plus_percent,
  higher_edu_18_plus_per_capita,
  pct_age_85_plus
)

# Adapted from http://www.sthda.com/english/articles/37-model-selection-essentials-in-r/155-best-subsets-regression-essentials-in-r/#computing-best-subsets-regression 
# (but without their use of BIC and CP):
models <- regsubsets(deathsperlakh~., data = data_cleaned_model_selection,force.in = "percent_in_poverty")
summary(models)

res.sum <- summary(models)
Adj.R2 <- which.max(res.sum$adjr2)
print(res.sum$adjr2)
cat('Optimal number of predictors to minimize adjusted R^2:', Adj.R2)
max_num_predictors <- length(res.sum$adjr2)

get_names <- coef(models, 1:sum(lengths(res.sum$adjr2)))

```


```{r}

# loop.vector <- sum(lengths(res.sum$adjr2))
#
# par_list <- list()
#
# for(i in 1:loop.vector){
#   name <- paste('model',i,sep='')
#   tmp <- list(names(get_names[[i]]))
#   tmp <- tmp[[1]][-1]
#  par_list[[name]] <- tmp}
#
#
# par_list

# adapted from http://www.sthda.com/english/articles/37-model-selection-essentials-in-r/155-best-subsets-regression-essentials-in-r/#k-fold-cross-validation :
get_model_formula <- function(id, object, outcome){
  # get models data
  models <- summary(object)$which[id,-1]
  # Get outcome variable
  #form <- as.formula(object$call[[2]])
  #outcome <- all.vars(form)[1]
  # Get model predictors
  predictors <- names(which(models == TRUE))
  predictors <- paste(predictors, collapse = "+")
  # Build model formula
  as.formula(paste0(outcome, "~", predictors))
}

outcomevariable <- "deathsperlakh"

model1 <- lm(as.formula(get_model_formula(1, models, outcomevariable)), data = data_cleaned_model_selection) 

model2 <- lm(as.formula(get_model_formula(2, models, outcomevariable)), data = data_cleaned_model_selection) 

model3 <- lm(as.formula(get_model_formula(3, models, outcomevariable)), data = data_cleaned_model_selection) 

model4 <- lm(as.formula(get_model_formula(4, models, outcomevariable)), data = data_cleaned_model_selection) 

model5 <- lm(as.formula(get_model_formula(5, models, outcomevariable)), data = data_cleaned_model_selection) 

model6 <- lm(as.formula(get_model_formula(6, models, outcomevariable)), data = data_cleaned_model_selection) 

if (max_num_predictors >= 7) {
model7 <- lm(as.formula(get_model_formula(7, models, outcomevariable)), data = data_cleaned_model_selection) 
}

if (max_num_predictors >= 8) {
  model8 <- lm(as.formula(get_model_formula(8, models, outcomevariable)), data = data_cleaned_model_selection) 
}

ourmodel <- lm(deathsperlakh~percent_in_poverty+
               pop_den+
               house_per_capita+
               hs_or_higher_25_plus_percent+
               pct_age_85_plus,
             data=data_cleaned_model_selection)
summary(ourmodel)

```

```{r}
ourmodel_minimal <- lm(deathsperlakh ~ percent_in_poverty, data = data_cleaned)
#stargazer(ourmodel_minimal, ourmodel, ourmodel2, ourmodel3, title="Model Comparison", align=TRUE,type='text')
```


### Check distributions and correlations
```{r}

pairs.panels(model.frame(model1))
pairs.panels(model.frame(model2))
pairs.panels(model.frame(model3))
pairs.panels(model.frame(model4))
pairs.panels(model.frame(model5))
pairs.panels(model.frame(model6))
if (max_num_predictors >= 7) {
pairs.panels(model.frame(model7))
}
if (max_num_predictors >= 8) {
pairs.panels(model.frame(model8))
}
pairs.panels(model.frame(ourmodel))

```



## Check CLM Assumptions


### IID?

_(no EDA needed, to be covered entirely in prose in the report)_


### Linear conditional expectation?

Eyeball the plot of residuals vs. the predicted death rate values: 

```{r}
data_cleaned %>%
  mutate(
    predicted_values = predict(ourmodel),
    residuals = resid(ourmodel)
    ) %>%
    ggplot(aes(predicted_values, residuals) ) +
    geom_point() +
    geom_text(aes(label = state), size=3) +
    stat_smooth() +
    labs(x="Predicted value", y="Residual", 
         title = "Residuals vs. predicted values (Model 2)")
```
```{r}
# equivalent plot:
plot(ourmodel, which = 1)
```

```{r}

p1 <- data_cleaned %>%
  mutate(
    predicted_values = predict(model1),
    residuals = resid(model1)
    ) %>%
    ggplot(aes(predicted_values, residuals) ) +
    geom_point() +
    stat_smooth() +
    labs(x="Predicted Value", y="Residual", title = "model1")


p2 <- data_cleaned %>%
  mutate(
    predicted_values = predict(model2),
    residuals = resid(model2)
    ) %>%
    ggplot(aes(predicted_values, residuals) ) +
    geom_point() +
    stat_smooth() +
    labs(x="Predicted Value", y="Residual", title = "model2")


p3 <- data_cleaned %>%
  mutate(
    predicted_values = predict(model3),
    residuals = resid(model3)
    ) %>%
    ggplot(aes(predicted_values, residuals) ) +
    geom_point() +
    stat_smooth() +
    labs(x="Predicted Value", y="Residual", title = "model3")


p4 <- data_cleaned %>%
  mutate(
    predicted_values = predict(model4),
    residuals = resid(model4)
    ) %>%
    ggplot(aes(predicted_values, residuals) ) +
    geom_point() +
    stat_smooth() +
    labs(x="Predicted Value", y="Residual", title = "model4")


p5 <- data_cleaned %>%
  mutate(
    predicted_values = predict(model5),
    residuals = resid(model5)
    ) %>%
    ggplot(aes(predicted_values, residuals) ) +
    geom_point() +
    stat_smooth() +
    labs(x="Predicted Value", y="Residual", title = "model5")


p6 <- data_cleaned %>%
  mutate(
    predicted_values = predict(model6),
    residuals = resid(model6)
    ) %>%
    ggplot(aes(predicted_values, residuals) ) +
    geom_point() +
    stat_smooth() +
    labs(x="Predicted Value", y="Residual", title = "model6")

if (max_num_predictors >= 7) {
p7 <- data_cleaned %>%
  mutate(
    predicted_values = predict(model7),
    residuals = resid(model7)
    ) %>%
    ggplot(aes(predicted_values, residuals) ) +
    geom_point() +
    stat_smooth() +
    labs(x="Predicted Value", y="Residual", title = "model7")
}

if (max_num_predictors >= 8) {
p8 <- data_cleaned %>%
  mutate(
    predicted_values = predict(model8),
    residuals = resid(model8)
    ) %>%
    ggplot(aes(predicted_values, residuals) ) +
    geom_point() +
    stat_smooth() +
    labs(x="Predicted Value", y="Residual", title = "model8")
}

if (max_num_predictors >= 7) {
  if (max_num_predictors >= 8) {
ggarrange(p1,p2,p3,p4,p5,p6,p7,p8, rremove("x.text"), 
          ncol = 2, nrow = 2)
  } else {
  ggarrange(p1,p2,p3,p4,p5,p6,p7, rremove("x.text"), 
          ncol = 2, nrow = 2)
}} else {
ggarrange(p1,p2,p3,p4,p5,p6, rremove("x.text"), 
          ncol = 2, nrow = 2)
}
```

### No perfect collinearity?

This assumption is satisfied if `lm()` did not drop coefficients:

```{r}
ourmodel$coefficients
```
We can also check correlations between variables:

```{r check correlations}
cor(select(data_cleaned, percent_in_poverty, house_per_capita))
```

### Homoskedastic errors?
1. check if the Breusch-Pagan test already rejects the null hypothesis that we 
have homoskedasticity:
```{r}
bptest(ourmodel)
```

```{r}
bptest(model1)
bptest(model2)
bptest(model3)
bptest(model4)
bptest(model5)
bptest(model6)
if (max_num_predictors >= 7) {
bptest(model7)
}
if (max_num_predictors >= 8) {
bptest(model8)
}
```


2. Again eyeball the first plot of residuals vs. predictions in the 
"Linear conditional expectation" section above, this time to check how much the 
height of the grey band varies

and/or

3. eyeball the scale-location plot, where the smoothed curve should look flat:

```{r}
plot(ourmodel, which = 3)
```

### Normally distributed errors?

1. check if the Shapiro-Wilk test already rejects the null hypothesis that the 
residuals are normally distributed:
```{r}
shapiro.test(resid(ourmodel))
```
and/or

2. eyeball the QQ-plot:
```{r}
plot(ourmodel, which=2)
```


```{r}

par(mfrow = c(2, 2))

plot(model1, which=2)
plot(model2, which=2)
plot(model3, which=2)
plot(model4, which=2)
plot(model5, which=2)
plot(model6, which=2)
if (max_num_predictors >= 7) {
plot(model7, which=2)
}
if (max_num_predictors >= 8) {
plot(model8, which=2)
}

```


# Other Graphs

```{r simple scatterplot,echo=FALSE,message=FALSE, fig.height=5, fig.width=9}
ggplot(data_cleaned, aes(x=pop_den, y=house_per_capita))+ 
  geom_point()+
  geom_text(aes(label = state),size=3)+
  labs(x = 'Population density (units: TKTK)',
    y = 'Inhabitants per housing units', 
    title = 'Housing unit size vs population density',
    subtitle = 'Data from ...',
    fontsize = 10)+
  theme(
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        plot.title = element_text(size=22))
# ggsave("TKTKT.png")
```

## Calculate cases on a time interval (from Kevin)
(Kevin:)
```{r calc_time_frame}
start.date <- as.Date("2021-01-01")
end.date <- as.Date("2021-02-28")
interval.cases <- NYT_Data[date == end.date, cases] - NYT_Data[date == start.date, cases]
```

## Total death ratio vs. % of aged population
```{r deaths_vs_aged_pc}
ggplot(data_cleaned, aes(x=pop_age_75_plus, y=deathsperlakh)) + 
  geom_point() + 
  geom_text(aes(label = state)) 
```

```{r simple deaths poverty scatterplot,echo=FALSE,message=FALSE, fig.height=5, fig.width=9}
ggplot(data_cleaned, aes(x=percent_in_poverty, y=deathsperlakh))+ 
  geom_point()+
  geom_text(aes(label = state),size=3)+
  labs(x = 'Percentage of Total Population in Poverty (2019)',
    y = 'Deaths per 100,000 People as of March 26, 2021', 
    title = 'Poverty and Covid Deaths',
    subtitle = 'Data from the 2019 American Community Survey and The New York Times Covid 19 Database',
    fontsize = 10)+
  theme(
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 12),
        plot.title = element_text(size=22))
#ggsave("Poverty_and_Covid_Deaths.png")
```